Improvements of Chatbot-Applications offering Cognitive Behavioural Therapy regarding Efficacy, Privacy and Safety

### 2021-06-05 paper.tex
improved references and typos

\emph{Gratitude Journaling} works by writing down moments, situations or facts the writer is grateful for. It can lead to a more positive mindset and higher mindfulness. 

@@

therapy adherence which can be seen as an important aspect of a successful therapy. 

@@

\section{Strenghts and weaknesses of existing chatbots}
\label{sec:two}
This far, there are a couple of chatbots available. This paper will mainly focus on the \emph{Woebot} Chatbot and \emph{Wysa} for several reasons.
On the one hand, they have been developed in a research oriented context. 
The examples are available in english language and free to use.\footnote{https://woebothealth.com/about-us/, last access: 01.05.2021}\textsuperscript{,}\footnote{https://www.wysa.io/faq, last access: 01.05.2021}
\\
Additionally, the developers provide numerous information on their website
like experience reports or technical aspects.
On the other hand, \emph{Wysa} and \emph{Woebot} have been considered by the most of the recent research.
They revealed some weaknesses of those chatbots, some chances and also risks concerning efficacy, privacy and safety.

@@

While also offering CBT like \emph{Woebot}, \emph{Wysa} also offers guided meditation, breathing and yoga. Additionally, it conducts \emph{Dialectial Behaviour Therapy} (DBT).\textsuperscript{7} 

@@

It is important to mention that the chatbot can not be used through third party applications like \emph{Facebook's} messenger. Hence a user can use the application more independently of privacy policies of other companies. Although it is not completely excluded as \emph{Wysa} allows to speak with the chatbot 

@@

As a consequence, governments around the globe define rules in order to protect the individual from exploitation and to ensure that users have access and control over personal data stored by service providers.
An example for it is the General Data Protection Regulation by the European Union. Considering chatbots for mental health, privacy is an important aspect. Sharing sensitive and personal information is part of the concept of CBT and hence necessary for a successful chatbot usage. Therefore users should constantly feel secure their data is being stored properly. 

@@

Firstly, it is necessary to define what privacy principles are concerning HCI. In this paper, the results of Patrick and Kenny \cite{Patrick2003} are used, who specifically derived privacy aspects by using the general EU legislation as basis.

@@

In order to summarize the privacy requirements of a chatbot for mental health it can be said that information is the most important one. Firstly, being informed about the processing of personal data leads to a conscious usage of the application which is essential to protect personal data. Secondly, information is necessary in order to know how data can be deleted or accessed. Finally it helps to better understand the system and 

@@

Donkin et. al \cite{Donkin} found a minimal completion rate of 56\% concerning online-therapy applications. Furthermore they revealed that the adherence correlates with the treatment outcomes.\cite{Donkin, Donkin2011}
\\This can be considered as a further proof for Route 1 in Figure \ref{fig:routes}.  

@@

They note that the way human beings think about abstract things like challenges or achievements is methaphorical by nature and experienced every day. Metaphors supporting mental health applications can strengthen those experiences.

@@

According to Ischen et.al \cite{Ischen}, services with a higher perception in anthropomorphism lead to a better adherence.

@@

This result corresponds to the results of Ludden et.al and hence supports the recommendation to implement the PERMA principles in applications for mental health like chatbots. 

@@

In order to respect the privacy and take user concerns serious, chatbot applications should only exist as a standlone app without the possibility to use it through a third party application such as \emph{Facebook}'s messenger in the long term. It is understandable that small companies with limited
financial possibilites tend to deploy their implementation via a third party application in order to reduce development costs. Moreover, the usage of well-established platforms like \emph{Facebook} helps to raise awareness on the one hand, while on the other hand it also reduces the effort for users since no installation is necessary. 

@@

work of Patrick and Kenny\cite{Patrick2003}. Third party applications can still be used as marketing platforms but the processing of sensitive data should be done within the control of the developing company.

@@

To summarize the idea of a knowledge and data dashboard it can be said that it has the potential to improve the trust of the users while simplyfing adminstration processes of the company at the same time. 

@@

The chatbot could also adapt its language to the age of the user, which could lead to a higher sense of well-being as the conversation resembles a dialogue with a friend. 

@@

While the consideration of the age can directly influence the questions asked during the conversation, aspects such as the design do not influence the dialog but could help to improve the acceptance of the application if the appearance adapts to user preferences. 

@@

be asked during the daily dialogs.

@@

Additionally to the Chatbot-Setup-Phase, too much ambient information should be avoided and the main functionality focused. An argument against it could be the funding of the app. The reviewed examples \emph{Woebot} and \emph{Wysa} do not advertise within the application but it is a conceivable approach which leads 

@@

As Grové \cite{Grove2021} points out, chabots for mental health should notice and react to critical situations in a sophisticated way due to the vulnerable target group. While the reviewed chatbots meet the requirements, Kretzschmar et. al \cite{Kretzschmar2019} defined,

@@

critical situations or what keywords lead to it in terms of natural language processing. Human therapists who consider to support their therapy with the help of a CBT-chatbot could easily assess its privacy capabilities with the provided information. This could lead to a higher use of chatbots in clinical settings.\\

@@

A further step in the development of this approach is the awarding of a privacy certificate for mental health chatbots by an independent company or institute. While the current supply is relatively low as there are only a few chatbots available, this approach would be worthwhile and purposeful if the supply and demand for the applications continue to grow. 
This certificate could include certain aspects which therapists and doctors consider as crucial for the privacy of a chatbot.\\ 
Additionally, results of current research such as the privacy aspects of Kretzschmar et. al \cite{Kretzschmar2019} that has been presented in \hyperref[sec:three_three]{Section 3.3} could be part of it.  
The certificate would then be considered as a quality mark for privacy.

@@

Thinking this further, the certificate could be expanded for privacy Issues and efficacy as well. It would turn it into a general quality feature for mental health chatbots. The certificate would enforce the implementation of certain aspects. While this can tackle the mentioned problems, it could also lead to the situation
that alternative concepts might not have a chance to compete on the market if the do not fulfil a certain aspect. Hence it would be necessary to keep the certificate dynamic. The conditions a chatbot application has to meet in order to get it should not be static.

\section{Summary}
\label{sec:five}
While the amount of people who use the internet as first source of information has increased\cite{Wyatt2015} it nowadays offers many applications for CBT such as chatbots. 
\\
Current and popular examples like \emph{Wysa} or \emph{Woebot} are available as a standalone app or through third party applications. Studies have shown the usage of them are efficiant, i.e they can improve the mood of their users significantly\cite{Fitzpatrick2017,Inkster}, but in order to tackle problems such as missing trust in a chatbot or a low therapy adherence, further improvements could be implemented like personalization 
or the provision of more transparency. To achieve that, a chatbot setup phase has been proposed in which specific questions could be asked concerning personal preferences and environment of the user. It could be considered as the preliminary talk prior to the therapy a human therapist conducts.
The chatbot setup phase would lead to an implementation of the PERMA aspects, that Ludden et. al \cite{Ludden2015} propose and which have been presented in \hyperref[sec:three_two]{Section 3.2}. Those aspects are promising to increase the general well-being of a user and hence the efficacy of the chabot application.
\\\\
The developing companies of both reviewed chatbots transparently explain what data is stored for what purpose.
Nevertheless, the review of general requirements for HCI applications has shown, that more information about data collection should be provided and how it is possible for the user to delete it\cite{Saglam2021}.
Furthermore, it has been found that users are not distinguishing between data and information\cite{Lee2011}.
In order to tackle those aspects, a knowledge and data dashboard is proposed that constantly provides information about the data the chatbot has collected and what it did derive from it.
The main advantages are the constant access to personal and sensitive data on the one hand. On the other hand, the provisioning of it could simplify and prevent administration processes such as deletion requests as they are done automatically via the dashboard.
\\\\
Considering privacy aspects, the reviewed chatbots already meet the requirements Kretzschmar et. al \cite{Kretzschmar2019} propose. Both of them explain that they are just robots and engage the user to do non-digital activities to prevent over-reliance.
It can be considered as a weakness that the only option to figure out if the user notifies and react to critical situations is a trial and error approach. Currently, the developing companies do not publish related information. In order to improve the transparency, and to ease the assessment if an application is suitable for therpeutical use, a certificate has been suggested, that would confirm the implementation of certain aspects and policies experts consider as crucial.
This certificate could be expaneded to privacy and efficacy features as well which could result in a general quality feature for mental health chatbots.

\section{Future Outlook}
In order to get an idea how the field of mental health chatbots will develop in the future, the product pipeline of \emph{WoebotHealth}\footnote{https://woebothealth.com/products-pipeline/, last access: 31.05.2021} gives an insight in their developments. They are moving forward to implement several chatbot solutions for specific target groups. Besides the presented general one for adult mental health, they also offer a variant of \emph{Woebot} for adolescents, pregnant persons or 
individuals with a problematic substance use.\textsuperscript{14}
\\
The approach to split up the general chatbot-application into several specific ones aiming different target groups could be promising to increase the efficacy even further due to therapy methods that are tailored to the patient's situation. 
\\\\
Furthermore \emph{WoebotHealth} is currently developing prescription digital therapeutics who aim to treat symptoms of certain mental illnesses. All of them are currently in the admission process, while the digital therapeutical \emph{WB001} has been designated by the \emph{FDA} (U.S Food and Drug Administration) as \emph{Breakthrough Device}.\textsuperscript{14}
\\
The development of prescription digital therapeutics could be one aspect regarding a more sustainable business model. Not only \emph{WoebotHealth} tries to reach health insurance providers, also \emph{Touchkin} announce on their website, that they are preparing to cooperate with them\footnote{https://www.wysa.io/, last access: 31.05.2021} what could increase the numbers of users.
Additionally they are also offering their services to employers who might have a genuine interest in the mental health of their employees. Using mental health application in the work place environment could play a bigger role in the near future as it is promising to fill the gap between meditation programs and employee assistance programs.\footnote{https://www.wysa.io/for-employers, last access: 31.05.2021}
\\\\
With the development of the internet of things, it is also conceivable that chatbots for mental health become ubiquitous. Being able to start a conversation constantly, for example with a smart watch, might have a positive impact on the therapy adherence. According to Donkin et. al \cite{Donkin2011}, this would also have a positive impact on the the efficacy of the application in general.
\\
Nevertheless, more research is necessary to examine the connection between ubiquitous computing and the use of mental health chatbots.
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{unsrt}
\bibliography{library}

>>>>

\emph{Gratitude Journaling} works by writing down moments, situations or facts the writer is grateful for. It can lead to a more positive mindset and higher mindfulness.\cite{Flinchbaugh2012} 

@@

therapy adherence which can be seen as an important aspect of a successful therapy.\cite{Suganuma2018} 

@@

\section{Strenghts and weaknesses of existing chatbots}
\label{sec:two}
This far, there are a couple of chatbots available. This paper will mainly focus on the \emph{Woebot} Chatbot and \emph{Wysa} for several reasons.
On the one hand, they have been developed in a research oriented context. 
The examples are available in english language and free to use.\footnote{https://woebothealth.com/about-us/, last access: 01.05.2021}\textsuperscript{,}\footnote{https://www.wysa.io/faq, last access: 01.05.2021}
\\
Additionally, the developers provide numerous information on their website
like experience reports or technical aspects.
On the other hand, \emph{Wysa} and \emph{Woebot} have been considered in most of recent research.
It revealed some weaknesses of those chatbots, some chances and also risks concerning efficacy, privacy and safety.

@@

While also offering CBT like \emph{Woebot}, \emph{Wysa} also offers guided meditation, breathing and yoga. Additionally, it conducts \emph{Dialectial Behaviour Therapy} (DBT).\footnote{https://emerj.com/ai-application-comparisons/chatbots-mental-health-therapy-comparing-5-current-apps-use-cases/, last access: 03.05.2021}

@@

It is important to mention that the chatbot can not be used through third-party applications like \emph{Facebook's} messenger. Hence a user can use the application more independently of privacy policies of other companies. Although it is not completely excluded as \emph{Wysa} allows to speak with the chatbot 

@@

As a consequence, laws and rules concerning the protection of personal data are evolving continiously\footnote{https://www.computerweekly.com/de/news/252498683/Wie-es-weltweit-um-den-Datenschutz-bestellt-ist, last access: 5.6.2021}.
An example for it is the General Data Protection Regulation by the European Union. Considering chatbots for mental health, privacy is an important aspect. Sharing sensitive and personal information is part of the concept of CBT and hence necessary for a successful chatbot usage. Therefore, users should constantly feel secure their data is being stored properly. 

@@

At first, it is necessary to define what privacy principles are concerning HCI. In this paper, the results of Patrick and Kenny \cite{Patrick2003} are used, who specifically derived privacy aspects by using the general EU legislation as basis.

@@

In order to summarize the privacy requirements of a chatbot for mental health it can be said that information is the most important one. Firstly, being informed about the processing of personal data leads to a conscious usage of the application which is essential to protect personal data. Secondly, information is necessary in order to know how data can be deleted or accessed. Finally, it helps to better understand the system and 

@@

Donkin et. al \cite{Donkin} found a minimal completion rate of 56\% concerning online-therapy applications. Furthermore they revealed in a meta review that the adherence correlates with the treatment outcomes positively\cite{Donkin2011}.
This can be considered as a further proof for Route 1 in Figure \ref{fig:routes}.  

@@

They note that the way human beings think about abstract things like challenges or achievements is methaphorical by nature and experienced every day. Metaphors supporting mental health applications could strengthen those experiences.

@@

According to Ischen et.al \cite{Ischen} and Suganuma et.al \cite{Suganuma2018}, services with a higher perception in anthropomorphism lead to a better adherence.

@@

This result corresponds to the results of Ludden et.al \cite{Ludden2015} and hence supports the recommendation to implement the PERMA principles in applications for mental health like chatbots. 

@@

In order to respect the privacy and take user concerns serious, chatbot applications should only exist as a standlone app without the possibility to use it through a third-party application such as \emph{Facebook}'s messenger in the long term. It is understandable that small companies with limited
financial possibilites tend to deploy their implementation via a third-party application in order to reduce development costs. Moreover, the usage of well-established platforms like \emph{Facebook} helps to raise awareness on the one hand, while on the other hand it also reduces the effort for users since no installation is necessary. 

@@

work of Patrick and Kenny\cite{Patrick2003}. Third-party applications can still be used as marketing platforms but the processing of sensitive data should be done within the control of the developing company.

@@

To summarize the idea of a knowledge and data dashboard it can be said that it has the potential to improve the trust of the users while simplifying adminstration processes of the company at the same time. 

@@

The chatbot could also adapt its language to the age of the user, which could lead to a higher sense of well-being as the conversation would resemble a dialogue with a friend. 

@@

While the consideration of the age can directly influence the questions asked during the conversation, aspects such as the design do not influence the dialogue but could help to improve the acceptance of the application if the appearance adapts to user preferences. 

@@

be asked during the daily dialogues.

@@

Additional to the Chatbot-Setup-Phase, too much ambient information should be avoided and the main functionality focused. An argument against it could be the funding of the app. The reviewed examples \emph{Woebot} and \emph{Wysa} do not advertise within the application but it is a conceivable approach which leads 

@@

As Grové \cite{Grove2021} points out, chabots for mental health should notice and react to critical situations in a sophisticated way due to the vulnerable target group. While the reviewed chatbots meet the requirements Kretzschmar et. al \cite{Kretzschmar2019} defined,

@@

critical situations or what keywords lead to it in terms of natural language processing. Human therapists considering to support their therapy with the help of a CBT-chatbot could easily assess its safety capabilities with the provided information. This could lead to a higher use of chatbots in clinical settings.\\

@@

A further step in the development of this approach is the awarding of a safety certificate for mental health chatbots by an independent company or institute. While the current supply is relatively low as there are only a few chatbots available, this approach would be worthwhile and purposeful if the supply and demand for the applications continue to grow. 
This certificate could include certain aspects which therapists and doctors consider as crucial for the safety of a chatbot.\\ 
Additionally, results of current research such as the safety aspects of Kretzschmar et. al \cite{Kretzschmar2019} that has been presented in \hyperref[sec:three_three]{Section 3.3} could be part of it.  
The certificate would then be considered as a quality mark for safety.

@@

Thinking this further, the certificate could be expanded for privacy issues and efficacy as well. It would turn it into a general quality feature for mental health chatbots. The certificate would enforce the implementation of certain aspects. While this can tackle the mentioned problems, it could also lead to the situation
that alternative concepts might not have a chance to compete on the market if they do not fulfil a certain aspect. Hence it would be necessary to keep the certificate dynamic. The conditions a chatbot application has to meet in order to get it should not be static.

\section{Summary}
\label{sec:five}
While the amount of people who use the internet as first source of information has increased\cite{Wyatt2015}, it nowadays offers many applications for CBT such as chatbots. 
\\
Current and popular examples like \emph{Wysa} or \emph{Woebot} are available as a standalone app or through third-party applications. Studies have shown the usage of them are efficiant, i.e they can improve the mood of their users significantly\cite{Fitzpatrick2017,Inkster}, but in order to tackle problems such as missing trust in a chatbot or a low therapy adherence, further improvements could be implemented like personalization 
or the provision of more transparency. To achieve that, a chatbot setup phase has been proposed in which specific questions could be asked concerning personal preferences and environment of the user. It could be considered as the preliminary talk prior to the therapy a human therapist conducts.
The chatbot setup phase would lead to an implementation of the PERMA aspects, that Ludden et. al \cite{Ludden2015} proposed and which have been presented in \hyperref[sec:three_two]{Section 3.2}. Those aspects are promising to increase the general well-being of a user and hence the efficacy of the chabot application.
\\\\
The developing companies of both reviewed chatbots transparently explain what data is stored for what purpose.
Nevertheless, the review of general requirements for HCI applications has shown, that more information about data collection should be provided and how it is possible for the user to delete it\cite{Saglam2021}.
Furthermore, it has been found that users are not distinguishing between data and information\cite{Lee2011}.
In order to tackle those aspects, a knowledge and data dashboard has been proposed that constantly provides information about the data the chatbot has collected and what it did derive from it.
The main advantages are the constant access to personal and sensitive data on the one hand. On the other hand, the provisioning of it could simplify and prevent administration processes such as deletion requests as they are done automatically via the dashboard.
\\\\
Considering safety aspects, the reviewed chatbots already meet the requirements Kretzschmar et. al \cite{Kretzschmar2019} proposed. Both of them explain that they are just robots and engage the user to do non-digital activities to prevent over-reliance.
It can be considered as a weakness that the only option to figure out if the chatbot notifies and react to critical situations is a trial and error approach. Currently, the developing companies do not publish related information. In order to improve the transparency and to ease the assessment if an application is suitable for therpeutical use, a certificate has been suggested that would confirm the implementation of certain aspects and policies experts consider as crucial.
This certificate could be expaneded to privacy and efficacy features as well which could result in a general quality feature for mental health chatbots.

\section{Future Outlook}
In order to get an idea how the field of mental health chatbots will develop in the future, the product pipeline of \emph{WoebotHealth}\footnote{https://woebothealth.com/products-pipeline/, last access: 31.05.2021} gives an insight in their developments. They are moving forward to implement several chatbot solutions for specific target groups. Besides the presented general one for adult mental health, they also offer a variant of \emph{Woebot} for adolescents, pregnant persons or 
individuals with a problematic substance use.
\\
The approach to split up the general chatbot-application into several specific ones aiming different target groups could be promising to increase the efficacy even further due to therapy methods that are tailored to the patient's situation. 
\\\\
Furthermore \emph{WoebotHealth} is currently developing prescription digital therapeutics who aim to treat symptoms of certain mental illnesses. All of them are currently in the admission process, while the digital therapeutical \emph{WB001} has been designated by the \emph{FDA} (U.S Food and Drug Administration) as \emph{Breakthrough Device}.\textsuperscript{15}
\\
The development of prescription digital therapeutics could be one aspect regarding a more sustainable business model. Not only \emph{WoebotHealth} tries to reach health insurance providers, also \emph{Touchkin} announce on their website\footnote{https://www.wysa.io/, last access: 31.05.2021} that they are preparing to cooperate with them what could increase the number of users.
\\
Additionally they are also offering their services to employers who might have a genuine interest in the mental health of their employees. Using mental health application in the work place environment could play a bigger role in the near future as it is promising to fill the gap between meditation and employee assistance programs.\footnote{https://www.wysa.io/for-employers, last access: 31.05.2021}
\\\\
With the development of the internet of things, it is also conceivable that chatbots for mental health become ubiquitous. Being able to start a conversation constantly, for example with a smart watch, might have a positive impact on the therapy adherence. According to Donkin et. al \cite{Donkin2011}, this would also have a positive impact on the the efficacy of the application in general.
\\
Nevertheless, more research is necessary to examine the connection between ubiquitous computing and the use of mental health chatbots.
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{unsrt}
\bibliography{library}
